---
title: "Singlecell GSEA Ranked Test"
author: "Mark J Lawson"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
          collapsed: true
          smooth_scroll: true
          print: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Performing a GSEA Rank Test on Single Cell Data

The following is a tutorial on how to perform a GeneSet Enrichmment Analysis (GSEA) rank test on single cell data using `Seurat`, `DESeq2`, and the `fGSEA` packages. In a nutshell, we use methods to determine differentially expressed genes (DEGs) and use their resulting statistical metrics to rank the genes for a GSEA test.

## Initializing the data

The following are the required packages.

```{r library, message=FALSE}
# packages for doing the analysis
library(Seurat) # For single cell analysis
library(SeuratData) # To get the example dataset
library(fgsea)  # Fast GSEA implementation
library(msigdbr)  # For gene sets
library(DESeq2) # For pseudo-bulk differential analysis
library(dplyr) # Tidy data manipulation
library(ggplot2) # Plotting tools
library(cowplot) # Combine plots
```

### Loading in the data

In the following code, we load in the `ifnb` dataset. In this experiment, PBMCs were split into a stimulated and control group and the stimulated group was treated with interferon beta. This dataset is annotated and can be pulled from the `SeuratData` package. Ultimately, we are going to see what pathways are being affected when we compare stimulated to control cells.

```{r load, message=FALSE, warning=FALSE}
# get data for doing analysis
InstallData("ifnb")

LoadData("ifnb") # load the data
# Update old Seurat object to accommodate new features
ifnb <- UpdateSeuratObject(ifnb)
```

### Get Genesets

Here we use the `msigdbr` package to pull in the Hallmark geneset. Other collections like KEGG/Reactome or GO can be pulled by adjusting the `collection` and `subcollection` parameters. Here we are pulling the genesets for humans, but mouse data can be pulled in either setting `species` to \"Mus musculus\" or `db_species` to \"MM\". The function `msigdbr_species()` lists the available options.

```{r genesets, message=FALSE, warning=FALSE}
# Get gene sets from MSigDB (e.g., Hallmark pathways)
gene_sets <- msigdbr(species = "Homo sapiens", collection = "H")
# Or use other collections:
# "C2" for curated gene sets (KEGG, Reactome, Wikipathways etc)
# "C5" for GO terms
# "C8" for cell type signatures

# Convert to list format for fgsea
pathways <- gene_sets %>%
  split(x = .$gene_symbol, f = .$gs_name)

```

## First approach: Simple comparison with FindMarkers

### Getting the DEGs
In situations where we do not have technical or biological replicates and thus are expecting no batch effects, we can split up the data by cell type and perform a comparison using the `Seurat` function `FindMarkers`. Ultimately we then have a list of differentially expressed genes and their corresponding metrics that can be fed into the GSEA rank test.

```{r findmarkers, message=FALSE, warning=FALSE}
Idents(ifnb) <- "seurat_annotations" # ensure that the idents are the cell types

# focus on one cell
ifnb_cd14 <- subset(ifnb, idents = "CD14 Mono")

# label cells with conditions
Idents(ifnb_cd14) <- "stim"
table(Idents(ifnb_cd14))

# find differentially expressed genes between markers
markers <- FindMarkers(
  ifnb_cd14,
  ident.1 = "STIM",
  ident.2 = "CTRL",
  test.use = "wilcox",  # or "MAST", "DESeq2"
  logfc.threshold = 0,  # Get all genes
  min.pct = 0.1
)
markers$gene_name <- rownames(markers) # make gene names a column in data frame

head(markers)
```
### Performing GSEA on the ranked list
In order to perform the analysis and to give a directionality to the pathway enrichment, `fGSEA` requires a value that indicates both the significance of a gene's differential expression and whether the gene is up-regulated or down-regulated. To do this, we take the `-log10` of the p-value and multiply it by the sign of the log2 fold change. Also, since we have p-values of 0, we are replacing those values with the smallest non-zero value, to avoid results of `Inf` due to the log10 of 0 being undefined.
The resulting `stat` value is then used to rank the genes and then passed to `fGSEA`.

```{r fgsea1, message=FALSE, warning=FALSE}
# find minimum non-zero p-value
min_nz_pval <- min(markers$p_val[markers$p_val >0])

# create "stat" value for ranking
markers$stat <- -log10(pmax(markers$p_val, min_nz_pval)) * sign(markers$avg_log2FC)

# Create ranked gene list (by log fold change or stat)
# Important: remove NAs and sort
ranked_genes <- markers %>%
  filter(!is.na(stat) &
           !is.infinite(stat)) %>%
  arrange(desc(stat)) %>%
  pull(stat, name = gene_name)

# Run GSEA
fgsea_results <- fgsea(
  pathways = pathways,
  stats = ranked_genes,
  minSize = 15, # Minimal size of a gene set to test
  maxSize = 500 # Maximal size of a gene set to test
)

# output results
fgsea_results %>%
  arrange(padj) %>%
  head(5)

```

### Plotting the results

#### Table Plot
```{r tableplot1, message=FALSE, warning=FALSE}
# Table plot of top pathways
topPathways <- fgsea_results %>%
  filter(padj < 0.05) %>%
  arrange(desc(abs(NES))) %>%
  pull(pathway)

plotGseaTable(
  pathways[topPathways],
  ranked_genes,
  fgsea_results,
  gseaParam = 0.5,
  pathwayLabelStyle = list(size=8)
)
```

#### Enrichment Plot

```{r enrichmentplot1, message=FALSE, warning=FALSE}
# Plot enrichment for top pathway
p1 <- plotEnrichment(
  pathways[["HALLMARK_INTERFERON_GAMMA_RESPONSE"]], 
  ranked_genes
) + labs(title = "IFN-γ Response")

p2 <- plotEnrichment(
  pathways[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]], 
  ranked_genes
) + labs(title = "Oxidative Phosphorylation")

plot_grid(p1, p2, ncol=1)
```

## Second approach: Pseudo-bulk analysis using DESeq2

### Getting the DEGs

```{r pseudo, message=FALSE, warning=FALSE}
# aggregate cells to create "pseudobulk" 
aggregate_ifnb <- AggregateExpression(ifnb, group.by = c("seurat_annotations", "stim"), 
                                      return.seurat = TRUE)
Idents(aggregate_ifnb) <- "seurat_annotations"
agg_ifnb_cdmono <- subset(aggregate_ifnb, idents=c("CD14 Mono",
                                                   "CD16 Mono"))

# create metadata for samples
sample_metadata <- agg_ifnb_cdmono@meta.data %>%
  select(stim, seurat_annotations, orig.ident) %>%
  distinct() %>%
  as.data.frame()
rownames(sample_metadata) <- sample_metadata$orig.ident

# create DESeq data set
dds <- DESeqDataSetFromMatrix(
  countData = agg_ifnb_cdmono[["RNA"]]$counts,
  colData = sample_metadata,
  design = ~ stim
)

# run DESeq
dds <- DESeq(dds)

# pull out results
res <- results(dds, contrast = c("stim", "STIM", "CTRL"))

# reformat for extracting
res <- as.data.frame(res)
res$gene_name <- rownames(res)
```

### Performing GSEA on the ranked list

```{r fgsea2, message=FALSE, warning=FALSE}
# Create ranked list
ranked_genes <- res %>%
  filter(!is.na(stat)) %>%
  arrange(desc(stat)) %>%
  pull(stat, name = gene_name)

# Run GSEA
fgsea_results <- fgsea(
  pathways = pathways,
  stats = ranked_genes,
  minSize = 15,
  maxSize = 500
)

# output results
fgsea_results %>%
  arrange(padj) %>%
  head(5)

fgsea_results %>%
  arrange(padj) %>%
  tail(5)

```

### Plotting the results

#### Table Plot

```{r tableplot2, message=FALSE, warning=FALSE}
# Table plot of top pathways
topPathways <- fgsea_results %>%
  filter(padj < 0.05) %>%
  arrange(desc(abs(NES))) %>%
  pull(pathway)

plotGseaTable(
  pathways[topPathways],
  ranked_genes,
  fgsea_results,
  gseaParam = 0.5,
  pathwayLabelStyle = list(size=8)
)

```

#### Enrichment Plot

```{r enrichmentplot2, message=FALSE, warning=FALSE}

# compare top pathway to bottom pathway
p1 <- plotEnrichment(
  pathways[["HALLMARK_INFLAMMATORY_RESPONSE"]], 
  ranked_genes
) + labs(title = "Inflammatory Response")

p2 <- plotEnrichment(
  pathways[["HALLMARK_NOTCH_SIGNALING"]], 
  ranked_genes
) + labs(title = "Notch Signaling")

plot_grid(p1, p2, ncol=1)
```

#### Session info
``` {r sessionn}
sessionInfo()
```



